---
title: "Exercise"
output:
  html_document:
    toc: false
    toc_float: false
---

The following exploratory analysis investigates the relationship between objectively measured exercise (steps, active minutes, calories) metrics and self-reported menstrual symptoms across different menstrual cycle phases. We use data from both study intervals (2022 and 2024) to examine whether physical activity is associated with symptom severity.

```{r, include = FALSE}
# Loading libraries
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(knitr)
library(kableExtra)
library(patchwork)
library(scales)

# Setting plot theme
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.discrete.colour = function(...) scale_colour_brewer(..., palette = "Pastel2"),
  ggplot2.discrete.fill   = function(...) scale_fill_brewer(..., palette = "Pastel2"),

  ggplot2.continuous.colour = function(...) scale_colour_distiller(..., palette = "PuRd"),
  ggplot2.continuous.fill   = function(...) scale_fill_distiller(..., palette = "PuRd")
)
```

### Data Summary

#### Data Quality Check
Five observations showed extreme step counts (≥50,000) likely due to device errors. These were excluded from step count visualizations only, all analyses use the complete dataset as extreme values were isolated to step counts.These observations are retained in all descriptive statistics and statistical models to preserve the complete dataset. They are excluded only from the step count distribution plot to prevent visual distortion.

```{r}
# Outlier check: Identifying and viewing extreme observations
extreme_obs = read.csv("./clean_data/all_daily_data.csv") |> 
  distinct(id, study_interval, day_in_study, .keep_all = TRUE) |> 
  filter(total_steps >= 50000 | # unlikely to complete more than 50000 steps a day
         total_active_min >= 1440 | # cannot have active exercise minutes for more than 24 hours
         total_calories >= 10000) |> # unlikely to burn more than 10000 calories in a day
  select(id, study_interval, day_in_study, 
         total_steps, total_active_min, total_calories) |> 
  arrange(desc(total_steps))

# Displaying extreme observations
  kable(extreme_obs,
        caption = "Outliers with Extreme Values",
        col.names = c("ID", "Year", "Day", "Steps", "Active Min", 
                     "Calories")) |> 
  kable_styling(bootstrap_options = c("striped", "hover"))
```

#### Descriptives: Exercise Metrics
```{r, message = FALSE, warning = FALSE}
# Loading data
all_daily_data = read.csv("./clean_data/all_daily_data.csv") |> 
  distinct(id, study_interval, day_in_study, .keep_all = TRUE)

# Exercise descriptives
overall_desc = all_daily_data |> 
  summarise(
    Variable = c("Daily Steps", "Active Minutes/Day", "Calories Burned", 
                 "Exercise Sessions/Day", "Days with Activity (%)"),
    Mean = round(c(mean(total_steps, na.rm = TRUE),
                   mean(total_active_min, na.rm = TRUE),
                   mean(total_calories, na.rm = TRUE),
                   mean(n_sessions, na.rm = TRUE),
                   NA), 2),
    SD = round(c(sd(total_steps, na.rm = TRUE),
                 sd(total_active_min, na.rm = TRUE),
                 sd(total_calories, na.rm = TRUE),
                 sd(n_sessions, na.rm = TRUE),
                 NA), 2),
    Median = round(c(median(total_steps, na.rm = TRUE),
                     median(total_active_min, na.rm = TRUE),
                     median(total_calories, na.rm = TRUE),
                     median(n_sessions, na.rm = TRUE),
                     NA), 2),
    IQR = round(c(IQR(total_steps, na.rm = TRUE),
                  IQR(total_active_min, na.rm = TRUE),
                  IQR(total_calories, na.rm = TRUE),
                  IQR(n_sessions, na.rm = TRUE),
                  NA), 2),
    Percent = round(c(NA, NA, NA, NA,
                      mean(exercised_flag > 0, na.rm = TRUE) * 100), 2)
  )

# Displaying table (and making NAs blank)
overall_desc |> 
  mutate(across(where(is.numeric), ~ifelse(is.na(.), "", .))) |> 
  kable(caption = "Descriptive Statistics: Exercise Metrics",
        align = c("l", "r", "r", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```
Overall, participants demonstrated moderate daily physical activity levels. The median daily step count was `r format(overall_desc$Median[1], big.mark = ",")` steps, with a mean of `r format(overall_desc$Mean[1], big.mark = ",")` steps (SD = `r format(overall_desc$SD[1], big.mark = ",")`). The large standard deviation and difference between mean and median indicate presence of extreme outlier observations (as identified earlier). Active exercise time averaged `r overall_desc$Mean[2]` minutes per day (Median = `r overall_desc$Median[2]` minutes), with participants burning an average of `r format(overall_desc$Mean[3], big.mark = ",")` calories per session (Median = `r format(overall_desc$Median[3], big.mark = ",")`). Most participants engaged in `r overall_desc$Mean[4]` exercise sessions per day (Median = `r overall_desc$Median[4]`). Physical activity was recorded on `r overall_desc$Percent[5]`% of study days, indicating high compliance with activity tracking.

```{r, message = FALSE, warning = FALSE}
# Comparison by study year
year_comp = all_daily_data |> 
  group_by(study_interval) |> 
  summarise(
    `N Days` = n(),
    `N Participants` = n_distinct(id),
    `Mean Steps` = round(mean(total_steps, na.rm = TRUE), 2),
    `SD Steps` = round(sd(total_steps, na.rm = TRUE), 2),
    `Mean Active Min` = round(mean(total_active_min, na.rm = TRUE), 2),
    `SD Active Min` = round(sd(total_active_min, na.rm = TRUE), 2),
    `% Days Active` = round(mean(exercised_flag, na.rm = TRUE) * 100, 2)
  ) |> 
  rename(Year = study_interval)

kable(year_comp,
      caption = "Exercise Metrics by Study Year",
      align = c("c", rep("r", 7))) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```
Exercise patterns differed substantially between study years. The 2024 cohort showed considerably higher activity levels compared to 2022 across all metrics. 2024 participants averaged `r year_comp$"Mean Active Min"[2]` active minutes per day compared to `r year_comp$"Mean Active Min"[1]` minutes in 2022. The proportion of days with recorded activity increased from `r year_comp$"% Days Active"[1]`% in 2022 to `r year_comp$"% Days Active"[2]`% in 2024.

#### Descriptives: Symptom Severity
```{r}
# Symptom descriptives
symptom_desc = all_daily_data |> 
  summarise(
    across(c(appetite_score, headaches_score, cramps_score, sorebreasts_score, fatigue_score, sleepissue_score, moodswing_score, foodcravings_score, indigestion_score, bloating_score),
           list(mean = ~mean(.x, na.rm = TRUE),
                sd = ~sd(.x, na.rm = TRUE),
                median = ~median(.x, na.rm = TRUE)),
           .names = "{.col}_{.fn}")
  ) |> 
  pivot_longer(everything(),
               names_to = c("symptom", ".value"),
               names_pattern = "(.+)_score_(.+)") |> 
  mutate(
    symptom = str_to_title(str_replace_all(symptom, "_", " ")),
    across(where(is.numeric), ~round(.x, 2))
  ) |> 
  # Alphabetical order
  arrange(symptom)

kable(symptom_desc,
      col.names = c("Symptom", "Mean", "SD", "Median"),
      caption = "Symptom Severity (Scale: 0 = None, 5 = Very High)",
      align = c("l", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```
Symptom severity varied across menstrual cycle phases and between individuals. Most physical symptoms showed low-to-moderate mean severity on the 0-5 scale: cramps (M = `r symptom_desc$mean[3]`, SD = `r symptom_desc$sd[3]`), fatigue (M = `r symptom_desc$mean[4]`, SD = `r symptom_desc$sd[4]`), and bloating (M = `r symptom_desc$mean[2]`, SD = `r symptom_desc$sd[2]`). Psychological symptoms including mood swings (M = `r symptom_desc$mean[8]`, SD = `r symptom_desc$sd[8]`) and appetite changes (M = `r symptom_desc$mean[1]`, SD = `r symptom_desc$sd[1]`) were similarly mild to moderate on average. The relatively large standard deviations across all symptoms reflect substantial variability in symptom experiences across the menstrual cycle.


### Distribution of Key Exercise Metrics
```{r, fig.width = 13, fig.height = 12}

exercise_analysis = all_daily_data

# Distribution of daily step count
p1 = ggplot(exercise_analysis |> filter(total_steps > 0, total_steps < 50000), 
             aes(x = total_steps, fill = factor(study_interval))) +
  geom_histogram(bins = 50, alpha = 0.7, position = "dodge") +
  geom_vline(xintercept = median(exercise_analysis$total_steps[exercise_analysis$total_steps > 0], na.rm = TRUE),
             linetype = "dashed", color = "black", linewidth = 1) +
  labs(title = "Daily Step Count Distribution",
       subtitle = paste("Median:", 
                       scales::comma(round(median(exercise_analysis$total_steps[exercise_analysis$total_steps > 0], na.rm = TRUE))), 
                       "steps"),
       x = "Total Daily Steps",
       y = "Count",
       fill = "Study Year") +
  scale_x_continuous(labels = scales::comma)

# Active minutes distribution
p2 = ggplot(exercise_analysis |> filter(total_active_min > 0), 
             aes(x = total_active_min, fill = factor(study_interval))) +
  geom_histogram(bins = 50, alpha = 0.7, position = "dodge") +
  geom_vline(xintercept = median(exercise_analysis$total_active_min[exercise_analysis$total_active_min > 0], na.rm = TRUE),
             linetype = "dashed", color = "black", linewidth = 1) +
  labs(title = "Active Minutes Distribution",
       subtitle = paste("Median:", 
                       round(median(exercise_analysis$total_active_min[exercise_analysis$total_active_min > 0], na.rm = TRUE), 1), 
                       "minutes"),
       x = "Total Active Minutes",
       y = "Count",
       fill = "Study Year")

# Calories Burned Distribution
p3 = ggplot(exercise_analysis |> filter(total_calories > 0), 
             aes(x = total_calories, fill = factor(study_interval))) +
  geom_histogram(bins = 50, alpha = 0.7, position = "dodge") +
  geom_vline(xintercept = median(exercise_analysis$total_calories[exercise_analysis$total_calories > 0], na.rm = TRUE),
             linetype = "dashed", color = "black", linewidth = 1) +
  labs(title = "Calories Burned Distribution",
       subtitle = paste("Median:", 
                       scales::comma(round(median(exercise_analysis$total_calories[exercise_analysis$total_calories > 0], na.rm = TRUE))), 
                       "calories"),
       x = "Total Calories Burned",
       y = "Count",
       fill = "Study Year") +
  scale_x_continuous(labels = scales::comma)

# Exercise Sessions Per Day
p4 = ggplot(exercise_analysis |> filter(n_sessions > 0), 
             aes(x = n_sessions, fill = factor(study_interval))) +
  geom_bar(alpha = 0.7, position = "dodge") +
  labs(title = "Exercise Sessions Per Day",
       x = "Number of Sessions",
       y = "Count",
       fill = "Study Year")

(p1 + p2) / (p3 + p4) + 
  plot_layout(guides = "collect")

```

**Step Count Distribution:** Daily steps on active days showed pronounced right-skew 
with a median of 4,455 steps. The bulk of 
observations fell between 2,000-8,000 steps, with a long tail extending beyond 20,000 steps. 
The 2024 distribution (orange) shifted notably rightward compared to 2022 (teal), indicating 
higher step counts. Five extreme observations (≥50,000 steps) were excluded from this 
visualization only, as they distorted the scale and were obvious data errors.

**Active Minutes Distribution:** Exercise duration exhibited similar right-skew 
(52 minutes), with most sessions lasting 20-80 minutes. 
The 2024 cohort showed longer session durations, with the distribution peak occurring at 
higher values than 2022. This pattern suggests not only more frequent exercise in 2024, 
but also longer duration per session.

**Calories Burned Distribution:** Energy expenditure followed the expected pattern given 
step and duration data (Median = 153 calories). The heavy concentration near zero with gradual right tail indicates predominantly 
light-to-moderate intensity exercise, with occasional high-intensity sessions. The 2024-2022 
difference parallels that observed in steps and active minutes duration.

**Exercise Sessions:** Most active days included 1-3 discrete exercise sessions, with a 
clear mode at 2 sessions. The 2024 cohort showed higher frequencies of 2+ sessions compared 
to 2022, suggesting either improved activity tracking or genuinely more structured exercise 
routines.

These distributions collectively demonstrate substantial right-skew, and high 
variability.


### Activity type distribution
```{r, fig.width = 10, fig.height = 17}
# Preparing activity types data
activities_long = exercise_analysis |> 
  filter(!is.na(activity_types), activity_types != "", activity_types != "NA") |> 
  select(id, study_interval, day_in_study, activity_types) |> 
  mutate(
    # Clean activity types column
    activity_types = str_replace_all(activity_types, "\\n", ","),
    activity_types = str_replace_all(activity_types, "\"", "")
  ) |> 
  # Split comma-separated activities into rows
  separate_rows(activity_types, sep = ",") |> 
  mutate(
    activity_types = str_trim(activity_types),
    activity_types = str_to_title(activity_types)
  ) |> 
  filter(activity_types != "", activity_types != "Na")

# Counting activities by year
activity_counts = activities_long |> 
  group_by(study_interval, activity_types) |> 
  summarise(n_days = n(), .groups = "drop") |> 
  group_by(study_interval) |> 
  mutate(pct = n_days / sum(n_days) * 100) |> 
  ungroup()

# Getting top 10 activities
top_activities = activity_counts |> 
  group_by(activity_types) |> 
  summarise(total = sum(n_days)) |> 
  arrange(desc(total)) |> 
  slice(1:10) |> 
  pull(activity_types)

activity_counts_top = activity_counts |> 
  filter(activity_types %in% top_activities)

# Plot: Activity type frequency by year
p1 = ggplot(activity_counts_top, 
             aes(x = reorder(activity_types, pct), y = pct, 
                 fill = factor(study_interval))) +
  geom_col(position = "dodge", alpha = 0.8) +
  coord_flip() +
  labs(title = "Most Common Activity Types by Study Year",
       x = "Activity Type",
       y = "% of Days with Exercise",
       fill = "Study Year") +
  theme_minimal() +
  theme(legend.position = "bottom")

p1
```
Walking dominated exercise routines in both study years, accounting for the vast majority of recorded activity sessions. Beyond walking, participants engaged in diverse activities including aerobic workouts, general sport, and strength training (weights). The 2024 cohort showed greater diversity in activity types, with more structured exercises such as aerobic workouts and weight training compared to 2022. Activities like outdoor biking, running, treadmill, and swimming appeared in both years but at much lower frequencies. Notably, some activities (e.g., weights, treadmill, swim, workout, hike) appeared exclusively or predominantly in 2024.

### Exercise patterns across menstrual cycle phases
```{r, fig.width = 10, fig.height = 10}

p1 = ggplot(exercise_analysis |> filter(!is.na(phase), total_steps > 0, total_steps < 50000), 
             aes(x = phase, y = total_steps, fill = phase)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.1, size = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "white") +
  facet_wrap(~study_interval, ncol = 2) +  # Side by side
  labs(title = "Daily Steps by Menstrual Cycle Phase",
       subtitle = "White diamonds represent mean values | Comparing study years",
       x = "Cycle Phase",
       y = "Total Daily Steps") +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))

p2 = ggplot(exercise_analysis |> filter(!is.na(phase), total_active_min > 0), 
             aes(x = phase, y = total_active_min, fill = phase)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.1, size = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "white") +
  facet_wrap(~study_interval, ncol = 2) +
  labs(title = "Active Minutes by Menstrual Cycle Phase",
       x = "Cycle Phase",
       y = "Total Active Minutes") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))

p1 / p2 + plot_layout(guides = "collect")
```
Physical activity varied across menstrual cycle phases, with patterns differing by year. In 2022, activity levels remained relatively stable across phases, with only slight differences between fertility, follicular, luteal, and menstrual phases. In contrast, 2024 showed more pronounced phase-related patterns, with the fertility phase demonstrating highest activity levels, followed by the follicular phase, and lower activity during luteal and menstrual phases. Both years exhibited substantial individual variability, as evident by the wide spread of data points and multiple outliers, suggesting that not all participants followed the same phase-related activity patterns. The 2024 cohort showed consistently higher activity across all phases compared to 2022, confirming the year differences observed in overall descriptive statistics. 

### Exercise vs. menstrual symptoms

```{r, fig.width = 10, fig.height = 10, warning = FALSE, message = FALSE}
symptoms_long = exercise_analysis |> 
  filter(total_active_min > 0) |> 
  select(id, total_active_min, appetite_score, headaches_score, sorebreasts_score, sleepissue_score, cramps_score, fatigue_score, 
         headaches_score, moodswing_score, foodcravings_score, indigestion_score, bloating_score) |> 
  pivot_longer(cols = ends_with("_score"), 
               names_to = "symptom", 
               values_to = "severity") |> 
  filter(!is.na(severity)) |> 
  mutate(symptom = case_when(
    symptom == "appetite_score" ~ "Appetite",
    symptom == "headaches_score" ~"Head Aches",
    symptom == "sorebreasts_score" ~ "Sore Breasts",
    symptom == "sleepissue_score" ~ "Sleep Issue",
    symptom == "cramps_score" ~ "Cramps",
    symptom == "fatigue_score" ~ "Fatigue",
    symptom == "headaches_score" ~ "Headaches",
    symptom == "moodswing_score" ~ "Mood Swings",
    symptom == "foodcravings_score" ~ "Food Cravings",
    symptom == "indigestion_score" ~ "Indigestion",
    symptom == "bloating_score" ~ "Bloating"
  ))

p4 = ggplot(symptoms_long, 
             aes(x = total_active_min, y = severity, color = symptom)) +
  geom_jitter(alpha = 0.2, size = 0.5, height = 0.1) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.5, alpha = 0.2) +
  facet_wrap(~symptom, ncol = 2) +
  labs(title = "Daily Activity vs Individual Symptoms",
       x = "Total Daily Activity (mins)",
       y = "Symptom Severity (0-5)") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(breaks = 0:5) +
  theme(legend.position = "none")

p4
```
The scatterplots examine crude associations between daily  activity (exercise) and each symptom:

**Negative slopes (potential benefits):**
- **Bloating**: Modest downward trend suggests higher activity associates with reduced bloating
- **Food cravings**: Slight negative association, though relationship appears weak
- **Indigestion**: Minimal negative slope with high variability

**Flat relationships (no clear association):**
- **Cramps**: Near-horizontal line despite being primary outcome
- **Appetite, headaches, sleep issues, sore breasts**: Essentially flat with wide confidence bands
- **Mood swings**: Slight positive but nearly flat, high uncertainty

**Positive slope (likely reverse causation):**
- **Fatigue**: Higher fatigue appears to reduce exercise, not vice versa

**Sleep issues** and **sore breasts** exhibited insufficient data spread to estimate reliable trends

Across all panels, several key patterns emerge. First, the wide confidence bands indicate substantial uncertainty in all bivariate relationships. Second, the considerable scatter reveals that individual-level associations vary widely, i.e., some participants show strong exercise-symptom relationships while others show none. Third, many data points cluster at low activity levels (0-100 minutes), with sparser observations at higher activity, reflecting the right-skewed activity distribution. Fourth, symptoms generally concentrate at lower severity levels (0-2 on 0-5 scale), with relatively few observations at high severity, consistent with the descriptive statistics.


# Modelling and inferential analysis

We used linear mixed-effects models to test associations between daily physical activity and menstrual symptom severity. This modeling approach was necessary because our data violate the independence assumption of standard linear regression. Mixed-effects models explicitly account for this hierarchical data structure through random effects.

For each of 11 symptoms, we fit models with:

-   **Outcome**: Symptom severity (0-5 scale)
-   **Predictor**: Total active exercise minutes per day
-   **Covariate**: Menstrual cycle phase (Menstrual, Follicular, Fertility, Luteal)
-   **Random effects**: Random intercepts for participants (1|id)

The random intercept structure accounts for repeated measures within individuals, including the 20 participants (47.6%) who contributed data in both 2022 and 2024 study intervals. Each participant's random intercept captures their baseline symptom tendency, allowing the model to distinguish between-person effects (do people who exercise more generally have fewer symptoms?) from within-person effects (on days when a person exercises more, do they experience fewer symptoms?). 

Models were fit using the lme4 package in R with REML = FALSE. Statistical significance was assessed at α = 0.05.
```{r}

# Model 1: Exercise → cramps
model_cramps = lmer(cramps_score ~ total_active_min + phase + (1|id), 
                     data = exercise_analysis,
                     REML = FALSE)
summary(model_cramps)

# Model 2: Exercise → fatigue
model_fatigue = lmer(fatigue_score ~ total_active_min + phase + (1|id), 
                     data = exercise_analysis,
                     REML = FALSE)
summary(model_fatigue)

# Model 3: Exercise → headaches
model_headaches = lmer(headaches_score ~ total_active_min + phase + (1|id), 
                     data = exercise_analysis,
                     REML = FALSE)
summary(model_headaches)

# Model 4: Exercise → mood swings
model_mood = lmer(moodswing_score ~ total_active_min + phase + (1|id), 
                     data = exercise_analysis,
                     REML = FALSE)
summary(model_mood)

# Model 5: Exercise → sleep issues
model_sleep = lmer(sleepissue_score ~ total_active_min + phase + (1|id), 
                     data = exercise_analysis,
                     REML = FALSE)
summary(model_sleep)

# Model 6: Exercise → bloating
model_bloating = lmer(bloating_score ~ total_active_min + phase + (1|id), 
                     data = exercise_analysis,
                     REML = FALSE)
summary(model_bloating)

# Model 7: Exercise → appetite
model_appetite = lmer(appetite_score ~ total_active_min + phase + (1|id), 
                     data = exercise_analysis,
                     REML = FALSE)
summary(model_appetite)

# Model 8: Exercise → sore breasts
model_sorebreasts = lmer(sorebreasts_score ~ total_active_min + phase + (1|id), 
                     data = exercise_analysis,
                     REML = FALSE)
summary(model_sorebreasts)

# Model 9: Exercise → food cravings
model_foodcravings = lmer(foodcravings_score ~ total_active_min + phase + (1|id), 
                     data = exercise_analysis,
                     REML = FALSE)
summary(model_foodcravings)

# Model 10: Exercise → indigestion
model_indigestion = lmer(indigestion_score ~ total_active_min + phase + (1|id), 
                     data = exercise_analysis,
                     REML = FALSE)
summary(model_indigestion)

```

```{r}
# Cleaning results:
# Extracting coefficients for exercise effect
results_summary = bind_rows(
  tidy(model_cramps, effects = "fixed") %>% mutate(outcome = "Cramps"),
  tidy(model_fatigue, effects = "fixed") %>% mutate(outcome = "Fatigue"),
  tidy(model_headaches, effects = "fixed") %>% mutate(outcome = "Head Aches"),
  tidy(model_mood, effects = "fixed") %>% mutate(outcome = "Mood Swings"),
  tidy(model_sleep, effects = "fixed") %>% mutate(outcome = "Sleep Issues"),
  tidy(model_bloating, effects = "fixed") %>% mutate(outcome = "Bloating"),
  tidy(model_appetite, effects = "fixed") %>% mutate(outcome = "Appetite"),
  tidy(model_indigestion, effects = "fixed") %>% mutate(outcome = "Indigestion"),
  tidy(model_sorebreasts, effects = "fixed") %>% mutate(outcome = "Sore breasts"),
  tidy(model_foodcravings, effects = "fixed") %>% mutate(outcome = "Food cravings")
) %>%
  filter(term == "total_active_min") %>%
  select(outcome, estimate, std.error, statistic, p.value) %>%
  mutate(
    significant = ifelse(p.value < 0.05, "Yes", "No"),
    p.value = format.pval(p.value, digits = 3, eps = 0.001)
  )

knitr::kable(results_summary)
```

Linear mixed-effects models showed limited evidence for associations between daily physical activity and menstrual symptom severity after controlling for cycle phase and individual differences. Of the 11 symptoms examined, only fatigue showed statistically significant relationships with exercise at α = 0.05: fatigue (β = -0.0026819	, SE = 0.0013608, p = 0.0493).

For fatigue, the negative coefficient indicates that each additional minute of daily active exercise was associated with a 0.0027 point decrease in fatigue severity on the 0-5 scale. 

The remaining 9 symptoms showed no significant associations with exercise. Effect estimates for these symptoms were near-zero with confidence intervals spanning positive and negative values, indicating no detectable relationship between daily exercise and symptom severity after controlling for menstrual cycle phase.

Interestingly, cramps, the primary outcome of interest given its prominence in menstrual symptom research, showed essentially no association with exercise (β = -0.00005, p = 0.9647). The point estimate had wide confidence intervals indicating  no evidence of either beneficial or harmful effects.


### Model Diagnostics

```{r model-diagnostics, fig.width=10, fig.height=4}

# Three essential plots
par(mfrow = c(1, 3))

# 1. Residuals vs Fitted (linearity + homoscedasticity)
plot(fitted(model_fatigue), resid(model_fatigue),
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted",
     pch = 16, col = rgb(0, 0, 0, 0.3))
abline(h = 0, col = "red", lty = 2, lwd = 2)

# 2. Q-Q Plot (normality of residuals)
qqnorm(resid(model_cramps), main = "Q-Q Plot: Residuals",
       pch = 16, col = rgb(0, 0, 0, 0.3))
qqline(resid(model_cramps), col = "red", lwd = 2)

# 3. Q-Q Plot (normality of random effects)
qqnorm(ranef(model_cramps)$id[,1], main = "Q-Q Plot: Random Effects",
       pch = 16, col = rgb(0, 0, 0, 0.5))
qqline(ranef(model_cramps)$id[,1], col = "red", lwd = 2)

par(mfrow = c(1, 1))
```
The residual vs fitted plot shows diagonal banding typical of ordinal outcome data (symptom scores are discrete: 0, 1, 2, 3, 4, 5). This is expected when applying linear models to Likert-scale data. The Q-Q plots indicate that residuals and random effects are approximately normally distributed, supporting the use of mixed-effects models.
