---
title: "Report"
output:
  html_document:
    code_folding: hide 
---

```{r, message=FALSE}
# Load packages 
library(tidyverse)
library(ggplot2)
library(modelr)
library(mgcv)
library(dplyr)
library(tidyr)
library(patchwork)
library(lme4)
library(lmerTest)
library(ggsci)

# Set plot theme
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.discrete.colour = function(...) scale_colour_brewer(..., palette = "Pastel2"),
  ggplot2.discrete.fill   = function(...) scale_fill_brewer(..., palette = "Pastel2"),

  ggplot2.continuous.colour = function(...) scale_colour_distiller(..., palette = "PuRd"),
  ggplot2.continuous.fill   = function(...) scale_fill_distiller(..., palette = "PuRd")
)
```

# Motivation
Half of the world's population has experineced or will experience menstruation; for many, this physiologic function occurs monthly for several decades. For young adults who menstruate, symptoms and regular menstration can be a vital health indicator. Abnormal symptoms can further suggest reproductive tract abnormalities, such as uterine fibroids or endometriosis. However, abnormal symptoms such as heavy bleeding or extreme pain can also be attributed to lifestyle characteristics. By leveraging a cohort of young adult menstruators, we hope to better understand the relationship between physiological menstrual characteristics and lifestyle factors such as stress, sleep, and exercise. Understanding how day-to-day factors influence menstrual health can help individuals who menstruate understand how to better manage concerning symptoms or identify when abnormal symptoms could be indicative of a more serious health concern. 

# Related Work

# Questions
1. What are background characteristics of people who participated in this study?
2. How does the menstrual cycle and hormones affect sleep and stress?
3. How does exercise and caloric consumption affect menstrual symptoms and hormonal levels? 
4. How do hormonal levels affect menstrual symptoms? 
5. What is the significance of menstruation on self-reported symptoms of participants in this study? 

# Data
## Data Sources

We utilized the inclusive, multimodal, longitudinal, and de-identified [mcPHASES](https://physionet.org/content/mcphases/1.0.0/) (menstrual cycle Physiological, Hormonal, and Self-Reported Events and Symptoms) dataset for menstrual health tracking with wearable devices.

Data was collected from 42 menstruating young adults in Canada over two 3-month intervals: Interval 1 (January–April 2022) and Interval 2 (July–October 2024). Participants wore Fitbit Sense smartwatches to measure physiological signals, and Mira Plus Starter Kits to track their hormone levels. Participants also self-reported daily symptom experiences like cramps, sleep quality, and stress levels via a smartphone diary app.

The original dataset contains 23 structured tables organized by signal category. For this project, we focused on the lifestyle factors such as exercise, sleep, and stress, and menstrual cycle symptoms. The datasets we used are the following:

-   Demographics:
    -   `subject_info.csv`: demographic data such as age, education, and age at menarche, and background survey responses collected at the start of the study
    -   `height_and_weight.csv`: participant's baseline height and weight
-   Hormone and Symptoms:
    -   `hormones_and_selfreport.csv`：hormone data (LH, E3G, PdG) from the Mira fertility device and daily self-reported symptom survey responses (e.g., cramps, mood, menstrual flow) on a 6-level Likert-type scale from 0 ("Not at all") to 5 ("Very high")
-   Lifestyle Factors:
    -   `stress_score.csv`: daily Stress Management Score from those who have access to the Stress Management experience
    -   `sleep.csv`: sleep session logs including timestamps, durations, and quality metrics collected by Fitbit
    -   `sleep_score.csv`: daily sleep score breakdowns provided by Fitbit, summarizing overall sleep quality based on multiple contributing factors

------------------------------------------------------------------------

**In total, we have 5 tidy dataframes: **

1.  `subject_info` 
    - Contains demographics and background survey responses data where each row represents a participant
    - A total of 62 observations of 18 variables
2.  `hormone_symptoms` 
    - Contains daily hormone and self-reported symptom-related data
    - A total of 5659 observations of 36 variables
3.  `sleep_stress_daily`
    - Contains daily sleep and stress related data
    - A total of 5400 observations of 7 variables 
4.  `exercise_daily` 
    - Contains daily exercise related data
    - A total of 1744 observations of 12 variables 
5.  `all_daily_data` 
    - The multimodal merged version of dataframes 2,3,4, where each row represents a **daily** data for a person in a specific study year, including hormone and self-reported symptoms, sleep score, stress score, daily exercise data.
    - A total of 5659 observations of 50 variables 

All dataframes can be linked by participant's `id`. Besides `subejct_info`, other dataframes can be further linked by `day_in_study` and `study_interval`.

------------------------------------------------------------------------

### `subject_info` - Subject-Relevant Data

The `subject_info` dataframe contains participants demographics data from `subject_info.csv` and `height_and_weight.csv`, as well as participants study enrollment year (this piece of information is from `hormones_and_selfreport.csv`).

It contains a total of 62 observations of 18 variables. Some key variables are the following:

-   `id`: participant's unique identifier
-   Age relevant variables: `birth_year` and `age_of_first_menarche`
-   Height and weight related variables: `avg_height`, `avg_weight`, `bmi`
-   Study enrollment related information: `study` (flag of which study the participant enrolled in, `1`=2022, `2`=2024, `3`=both), `year` (the year of study)
-   Categorical/Factor variables like `ethnicity`, `education`, `self_report_menstrual_health_literacy`, etc.

```{r, message=FALSE}
ht_wt = read_csv("data/height_and_weight.csv")|> 
  janitor::clean_names() |> 
  # calculate mean height/weight
  rowwise() |>
  mutate(
    avg_height = mean(c(height_2022, height_2024), na.rm = TRUE),
    avg_weight = mean(c(weight_2022, weight_2024), na.rm = TRUE)
  ) |>
  ungroup()

# import csv file
hormones_and_selfreport = read_csv("data/hormones_and_selfreport.csv") |> 
  janitor::clean_names() 
# extract study_flag
study_flag <- hormones_and_selfreport  |> 
  distinct(id, study_interval) |> 
  mutate(flag = 1,
         study_interval = paste0("study_", study_interval)) |> 
  pivot_wider(
    names_from = study_interval,
    values_from = flag,
    values_fill = 0  
  )

# read in subject-info.csv
subject_info_raw = read_csv("data/subject-info.csv") |> 
  janitor::clean_names()

subject_info <- subject_info_raw  |> 
  # merging
  left_join(ht_wt, by = "id") |> 
  left_join(study_flag, by = "id") |>
  # add a study_flag
  mutate(
    study = case_when(
      study_2022 == 1 & study_2024 == 1 ~ 3L,  # both 2022 & 2024
      study_2022 == 1 & study_2024 == 0 ~ 1L,  
      study_2024 == 1 & study_2022 == 0 ~ 2L,  
      TRUE ~ NA_integer_
    )
  ) |> 
  # pivot_long for wt & ht
  pivot_longer(
    cols = c(weight_2022, weight_2024, height_2022, height_2024),
    names_to  = c(".value", "year"), 
    names_sep = "_"                 
  ) |> 
  rename(
    weight_yr = weight,
    height_yr = height
  ) |> 
  # keep only the relevant study year
  filter(
    (study == 1 & year == 2022) |
    (study == 2 & year == 2024) |
    (study == 3)
  ) |> 
  # calculate age
  mutate(
    year = as.numeric(year),
    birth_year = as.numeric(birth_year),
    age = year - birth_year
  ) |> 
  # get a mean age for each id
  group_by(id) |> 
  mutate(
    mean_age = mean(age, na.rm = TRUE)
  ) |> 
  ungroup() |> 
  # get BMI
  mutate(
    bmi = weight_yr / ( (height_yr / 100)^2 )
  )|>
  # convert NaN to NA
  mutate(
    across(c(avg_weight, avg_height), 
           ~ if_else(is.nan(.), NA_real_, .))
  ) |> 
  #this factors variables and also creates "gender abrv" and "education abrv", which are simplified versions of these variables for cleaner tables and figures 
  mutate(
    education = str_trim(education),
    education_abrv = case_when(
      education == "High school degree or equivalent (e.g. GED)" ~ "High School",
      education == "Some university/ post-secondary, no degree" ~ "Some College",
      education == "Bachelor's degree (e.g. BA, BS)" ~ "Bachelors",
      education %in% c("Master's degree (e.g. MA, MS, MEd)", "Doctorate or professional degree") ~ "Graduate Degree or Higher",
      TRUE ~ "Other"),
    education_abrv = factor(
      education_abrv,
      levels = c("High School", "Some College", "Bachelors", "Graduate Degree or Higher", "Other"),
      ordered = TRUE),
     self_report_menstrual_health_literacy = factor(
      self_report_menstrual_health_literacy,
      levels = c("Non-existant", "Low", "Medium", "High"),
      ordered = TRUE),
    gender_abrv = case_when(
      gender %in% c("Gender Fluid", "Non-binary") ~ "Gender Fluid or Non-binary", 
      gender %in% c("Prefer not to say", "Other") ~ "Prefer not to say or Other", 
      gender == "Woman" ~ "Woman"),
      gender_abrv = factor(gender_abrv, levels = c("Woman", "Gender Fluid or Non-binary", "Prefer not to say or Other")), 
      sexually_active = factor(sexually_active, levels = c("Yes", "No", "Prefer not to say"))) |> 
  # only need 1 study_interval flag to indicate the study year
  mutate(study_interval = year) |> 
  select(-study_2022, -study_2024, -year, -study) |> 
  # rearrange the column order
  select(
    id,     
    study_interval,     
    everything()
  )
  
```

### `hormone_symptoms` - Daily Hormone and Self-Reported Symptom-Related Data

The `hormone_symptoms` dataframe contains participants' daily hormone data (LH, E3G, PdG) and self-reported symptom-related data (e.g., cramps, mood, menstrual flow) on a 6-level Likert-type scale from 0 ("Not at all") to 5 ("Very high"), from `hormones_and_selfreport.csv`.

It contains a total of 5659 observations of 36 variables. Some key variables are the following:

-   `id`: participant's unique identifier
-   `study_interval`: the year of study that participant enrolled in 2022 or 2024)
-   `day_in_study`: normalized day index starting from day 1 of each participant’s study interval
-   `phase`: factor menstrual cycle phase label (follicular, ovulation, luteal, or fertility)
-   Hormone data: `lh` (luteinizing hormone (mIU/mL) level, `estrogen` (estrone-3-glucuronide level (ng/mL)), and `pdg` (pregnanediol glucuronide (mcg/mL))
-   Factor flow data: `flow_volume` on a Likert scale (Lowest = “Not at all” to highest = “Very Heavy”), and `flow_color` (categorical indicator of menstrual flow or discharge color)
-   Factor symptoms severity data on Likert scale, such as `cramps`, `fatigue`, and `moodswing`.
-   Numeric symptoms severity data converted from factor symptoms severity data, such as `cramps_score`, `fatigue_score`, and `moodswing_score`.

```{r, message=FALSE}
# read in raw csv file
hormone_symptoms_raw <- 
  read_csv("data/hormones_and_selfreport.csv") |> 
  janitor::clean_names() 

# factor string values
likert_level_factor = c("not at all", "very low/little", "low", "moderate", "high", "very high")
likert_levels_no_verylowlittle = c("not at all", "very low", "low", "moderate", "high", "very high")
# likert_scale for numeric score conversion
likert_6_levels = c("not at all" = 0, "very low/little" = 1, "very low" = 1, "low" = 2, "moderate" = 3, "high" = 4, "very high" = 5)


hormone_symptoms <- hormone_symptoms_raw |> 
  # convert all to lower case string for standardization
  mutate(across(flow_volume:bloating, 
                ~ ifelse(!is.na(.) & !grepl("^\\d+$", .), tolower(.), .))) |> 
  mutate(across(c(phase), ~ ifelse(!is.na(.) & !grepl("^\\d+$", .), tolower(.), .))) |> 
  # factor variables
  mutate(
    phase = factor(phase, levels = c("menstrual", "follicular", "fertility", "luteal")),
    flow_volume = factor(flow_volume, 
                         levels = c("not at all", "spotting / very light","light","moderate", "somewhat heavy","heavy", "very heavy")),
    flow_color = factor(flow_color,
                        levels = c("not at all", "pink","bright red", "dark brown / dark red", "grey", "black","other")),
  ) |> 
  # factor character values
  mutate(
    across(
    .cols = c(sleepissue, headaches, stress, cramps, foodcravings, indigestion, bloating, moodswing, sorebreasts, fatigue),  
    .fns = ~ factor(., levels = likert_level_factor),  
    .names = "{.col}"  
  )
  ) |> 
  mutate(across(
    .cols = c(appetite, exerciselevel),
    .fns = ~ factor(., levels = likert_levels_no_verylowlittle), 
    .names = "{.col}"  
  )) |> 
  # convert categorical variable values to numeric intensity scores
  mutate(across(
    .cols = appetite:bloating,
    .fns = ~ recode(., !!!likert_6_levels),
    .names = "{.col}_score"
  )) |> 
  # special conversion for flow
  mutate(
    flow_volume_score = case_when(
      flow_volume == "not at all" ~ 0,
      flow_volume == "spotting / very light" ~ 1,
      flow_volume == "light" ~ 2,
      flow_volume == "moderate" ~ 3,
      flow_volume == "somewhat heavy" ~ 4,
      flow_volume == "heavy" ~ 5,
      flow_volume == "very heavy" ~ 6,
      TRUE ~ NA_real_
    ),
    flow_color_score = case_when(
      flow_color == "not at all" ~ 0,
      flow_color == "pink" ~ 1,
      flow_color == "bright red" ~ 2,
      flow_color == "dark brown / dark red" ~ 3,
      flow_color == "grey" ~ 4,
      flow_color == "black" ~ 5,
      flow_color == "other" ~ 6,
      TRUE ~ NA_real_
    )
  )
```

### `sleep_stress_daily` - Daily Sleep and Stress Related Data

The `sleep_stress_daily` dataframe contains participants' daily sleep and stress data from 3 csv files (`sleep.csv`, `sleep_score.csv` and `stress_score.csv`).

It contains a total of 5400 observations of 7 variables. Some key variables are the following:

-   `id`: participant's unique identifier
-   `study_interval`: the year of study that participant enrolled in 2022 or 2024)
-   `day_in_study`: normalized day index starting from day 1 of each participant’s study interval
-   Sleep session logs data like `timeinbed_mean` (average minutes in bed) and `deep_sleep_in_minutes` (minutes spent in deep sleep)
-   `sleep_score`: overall sleep score out of 100, the higher the better
- `stress_score`: overall sleep score out of 100, the higher the better

```{r, load-sleep-stress-scores-data, message=FALSE, results='hide'}
# load raw data from csv files
sleep_score_raw <- read_csv("data/sleep_score.csv") |>
  janitor::clean_names()
sleep_raw       <- read_csv("data/sleep.csv") |>
  janitor::clean_names()
stress_score_raw <- read_csv("data/stress_score.csv") |>
  janitor::clean_names()

# combine the sleep and stress score
sleep_score <- sleep_score_raw |>
  select(id,study_interval, day_in_study, overall_score, deep_sleep_in_minutes) |> 
  mutate(sleep_score = overall_score) |> 
  select(-overall_score) |> 
  # remove duplicates by keeping the maximum score
  group_by(id, study_interval, day_in_study) |> 
  slice_max(order_by = sleep_score, n = 1, with_ties = FALSE) |> 
  ungroup()

stress_score <- stress_score_raw |>
  filter(calculation_failed == "FALSE")|>
  select(id, study_interval, day_in_study, stress_score)|>
  distinct(id, study_interval, day_in_study, .keep_all = TRUE)

sleep_stress_score <- full_join(sleep_score, stress_score, by=c("id", "study_interval", "day_in_study"))

# working on sleep data
sleep <- sleep_raw|>
  select(id,study_interval, sleep_end_day_in_study, timeinbed, mainsleep) |>
  # changed to sleep_end_day
  mutate(day_in_study = sleep_end_day_in_study) |>
  select(-sleep_end_day_in_study) |> 
  # only keep main sleep = TRUE as mainsleep: boolean indicating whether this is the main sleep session of the day
  filter(
    mainsleep == TRUE
  ) |> 
# check duplicates >> all duplicates from 2024
  group_by(id, study_interval, day_in_study) |>
  summarise(
    timeinbed_mean = mean(timeinbed, na.rm = TRUE),
    mainsleep = first(mainsleep),
    .groups = "drop" 
  ) |>
  ungroup() |> 
  # no need flag
  select(-mainsleep)

# MERGE
sleep_stress_daily <- full_join(sleep, sleep_stress_score, by=c("id", "study_interval", "day_in_study"))
```

### `exercise_daily` - Daily Exercise Related Data

The `exercise_daily` dataframe contains participants' daily exercise related data from `exercise.csv`).

It contains a total of 1744 observations of 12 variables. Some key variables are the following:

-   `id`: participant's unique identifier
-   `study_interval`: the year of study that participant enrolled in 2022 or 2024)
-   `day_in_study`: normalized day index starting from day 1 of each participant’s study interval
-   Exercise session logs data:
    - Numeric data like `total_duration_min` (amount of time spent in active movement during the session in minutes), `total_steps` (total number of steps recorded), `total_calores` (total calories burned), heart rate data (`avg_heartrate` and `max_heartrate`)
    - Categorical data like `activity_types` (e.g. sport, walk)
- `n_sessions`: numeric indicator of how many exercise sessions a participant had per day

```{r, message = FALSE}
# Read data
exercise_raw <- read_csv("data/exercise.csv") |> 
  janitor::clean_names() 

exercise_daily <- exercise_raw |> 
  distinct() |> 
  # aggregate to daily level
  group_by(id, study_interval, start_day_in_study) %>%
  summarise(
    total_calories = sum(calories, na.rm = TRUE),
    total_duration_min = sum(duration, na.rm = TRUE) / 60000,
    total_active_min = sum(activeduration, na.rm = TRUE) / 60000,
    total_steps = sum(steps, na.rm = TRUE),
    avg_heartrate = mean(averageheartrate, na.rm = TRUE),
    max_heartrate = ifelse(all(is.na(averageheartrate)), NA, max(averageheartrate, na.rm = TRUE)),  # Handling -Inf
    total_elevation = sum(elevationgain, na.rm = TRUE),
    n_sessions = n(),
    activity_types = paste(unique(activityname), collapse = ", "),
    .groups = "drop"
  ) |> 
  rename(day_in_study = start_day_in_study) 
```



### `all_daily_data` - Merged Multimodal Daily Data

The `all_daily_data` is a merged multimodal dataframe that contains all **daily-level** data from `hormone_symptoms`, `exercise_daily`, and `sleep_stress_daily` dataframes, and used `hormone_symptoms` as the anchor dataframe. Each row represents a daily data for a person in a specific study year (including sleep score, stress score, daily exercise data, hormone and self-reported symptoms).

It contains a total of 5659 observations of 50 variables, with unique combination of key identifiers (`id`, `study_interval`, and `day_in_study`). Additional flag variables that are not in 3 multimodal dataframes are `exercise_flag` and `sleep_stress_flag` to indicate if the observation day had at least one recorded exercise session or if both stress and sleep scores are present.
```{r, message=FALSE}
all_daily_data <- hormone_symptoms |> 
  # MERGE exercise & hormone
  left_join(exercise_daily, by = c("id", "study_interval", "day_in_study")) |>
  # add flag exercised_flag for easy filtering in the future analyses
  # only need to keep exercised_flag==1 to have all rows with exercise data
  mutate(
    exercised_flag = ifelse(!is.na(n_sessions) & n_sessions > 0, 1, 0)
  ) |> 
  # MERGE sleep & stress score data
  # noticed hormone_symptoms df has a stress_score column 
  # drop it
  select(-stress_score) |> 
  left_join(sleep_stress_daily, by = c("id", "study_interval", "day_in_study")) |> 
  # add flag sleep_stress_flag for easy filtering in the future analyses
  mutate(
    sleep_stress_flag = ifelse(
      !is.na(stress_score) & !is.na(sleep_score),1,0
    )
  )  
  # MERGE demographic data
  # left_join(subject_info, by = c("id", "study_interval")) 
```



# Exploratory Data Analysis

## Participant Characteristics
### Age distribution
- View mean age in total (2022 & 2024)
```{r}
overall_n <- subject_info |> 
  distinct(id) |> 
  count(name = "n")

age_boxplot_all <- subject_info |> 
  ggplot(aes(x="All participants", y = mean_age)) +
  geom_boxplot(fill = "#bcbddc")+
  geom_text(
    data = overall_n,
    aes(
      x = "All participants", 
      y = max(subject_info$age, na.rm = TRUE) + 3,
      label = paste0("n = ", n)
    )
  ) +
  labs(
    x = NULL,
    y = "Mean age (years)",
    title = "Mean age of all participants"
  ) +
  ylim(15, NA) +
  theme(
    axis.text.x = element_text(size = 10),
    plot.title  = element_text(hjust = 0.5)
  )
age_boxplot_all
```
The cohort includes 42 participants with a median age of ~21 years and a small number of older outliers (ages 27 and 29). This relatively narrow, young age range reduces age-related heterogeneity in reproductive hormones and symptomatic expression, helping to isolate phase-related effects. 

### Age of first menarche
```{r}
menarche_plot = ggplot(subject_info, aes(x = "", y = age_of_first_menarche)) +
  geom_boxplot(fill = "cornflowerblue") +
  labs(title = "Age of Menarche",
    y = "Age (years)",
    x = "") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_blank(),
    axis.ticks.x = element_blank())

menarche_plot
```
Age of first menarche ranges roughly from 10 to 15 years with a median around 12 and a tight IQR (≈11–12.75). This homogeneous menarche distribution suggests a relatively standard reproductive development history across participants; because very early or very late menarche can influence cycle regularity and hormonal profiles, the lack of extreme variability here supports attributing observed hormone–symptom patterns primarily to current cycle phase rather than developmental differences.

### BMI

```{r bmi-boxplot}
overall_bmi_n <- subject_info |> 
  filter(!is.na(bmi)) |> 
  summarize(n = n())

bmi_all <- subject_info |> 
  filter(!is.na(bmi)) |> 
  ggplot(aes(x = "All participants", y = bmi)) +
  geom_boxplot(fill = "#bcbddc") +
  geom_text(
    data = overall_bmi_n,
    aes(
      x = "All participants",
      y = 45, 
      label = paste0("n = ", n)
    )
  ) +
  labs(
    x = NULL,
    y = "BMI (kg/m^2)",
    title = "Overall Mean BMI Distribution"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(size = 10)
  )

bmi_all
```
Among the 26 participants with BMI data, the median BMI ≈ 22.5 kg/m² (IQR ≈ 20–26) with one high outlier (~39). Overall, most participants fall in the normal-to-overweight range. Because adiposity influences sex steroid metabolism and inflammation, it is important to note that BMI variability could contribute to between-person differences in both hormone levels and symptom severity; including BMI as a covariate in later models may help separate phase effects from adiposity-driven variation. 


## Hormone

```{r, include=FALSE}
hormone_symptoms_eda=
  hormone_symptoms|>
  filter(!is.na(phase) & !is.na(moodswing) & !is.na(fatigue)& !is.na(flow_color)& !is.na(flow_volume)& !is.na(appetite))
```

LH & Estrogen
We want to examine average estrogen and LH levels across menstrual cycle phases to see if hormonal patterns match biological expectations
```{r, echo=FALSE}
hormone_avg <- hormone_symptoms_eda |>
  group_by(phase) |>
  summarise(
    estrogen_mean = mean(estrogen, na.rm = TRUE),
    lh_mean = mean(lh, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(hormone_avg, aes(x = phase)) +
  geom_line(aes(y = estrogen_mean, color = "Estrogen", group = 1), size = 1) +
  geom_line(aes(y = lh_mean, color = "LH", group = 1), size = 1) +
  geom_point(aes(y = estrogen_mean, color = "Estrogen"), size = 3) +
  geom_point(aes(y = lh_mean, color = "LH"), size = 3) +
  labs(title = "Average Estrogen and LH Across Cycle",
       x = "Phase", y = "Hormone Level", color = "Hormone") +
  theme_minimal()+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    plot.title = element_text(face = "bold"),
    panel.grid.major = element_line(color = "grey90")
  )
```
Estrogen levels are highest in the Fertility (ovulatory) phase, second highest in Luteal, and relatively low in Menstrual, gradually rising from Follicular to Luteal. LH is much lower than estrogen overall, peaking sharply in Fertility and remaining low in the other phases.

## Life factors
### Mood Swing

Moodswing Distribution Across Phases

Using violin plots to assess whether menstrual cycle hases influence mood swing intensity. 
```{r, echo=FALSE}
# Mood swing
ggplot(hormone_symptoms_eda, aes(x = phase, y = moodswing_score)) +
  geom_violin(trim = TRUE, fill = "orchid", alpha = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
  scale_y_continuous(limits = c(0,5), breaks = 0:5) +
  labs(
    title = "Distribution of Mood Swing Intensity Across Phases",
    x = "Phase",
    y = "Mood Swing (0–5)"
  ) + 
  theme_minimal()+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    plot.title = element_text(face = "bold"),
    panel.grid.major = element_line(color = "grey90")
  )

```
Violin plots show substantial inter-individual variability. Menstrual phase has the highest mood swing intensity (right-skewed distribution, peak around 1.85), while Luteal phase shows lower values with a wider base (more individuals with low mood swing).


Average Mood Swing Across Phases
To visualize general trends, we should summarize the average mood swing intensity across phases.
```{r, echo=FALSE}
symptom_means <- hormone_symptoms_eda |>
  group_by(phase) |>
  summarise(
    avg_moodswing = mean(moodswing_score, na.rm = TRUE),
    .groups = "drop"
  )

# Mood swing single line plot
ggplot(symptom_means, aes(x = phase, y = avg_moodswing, group = 1)) +
  geom_line(color = "purple", size = 1.5, alpha = 0.7) +
  geom_point(color = "purple", size = 3) +
  geom_text(aes(label = round(avg_moodswing, 2)), vjust = -1, size = 4, fontface = "bold") +
  labs(
    title = "Average Mood Swing Intensity Across Menstrual Cycle Phases",
    subtitle = "Higher values indicate more severe mood swings (Scale: 1-4)",
    x = "Phase",
    y = "Average Mood Swing Intensity"
  ) +
  ylim(0, 4) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    plot.title = element_text(face = "bold"),
    panel.grid.major = element_line(color = "grey90")
  )

```

Average mood swing is highest in Menstrual phase (1.85), lowest in Luteal phase (1.41), with intermediate levels in follicular and fertility phases.



### Flow Volume

Flow Volume Distribution Across Phases

To assess flow volume distribution patterns, we visualize flow volume variation across phases with violin plots.
```{r, echo=FALSE}
#violin
ggplot(hormone_symptoms_eda, aes(x = phase, y = flow_volume_score)) +
  geom_violin(fill = "skyblue", alpha = 0.5) +  
  labs(
    title = "Distribution of Flow Volume Across Phases",
    x = "Phase", y = "Flow Volume (0–6)"
  ) +
  theme_minimal()
```
Luteal phase shows a right-skewed distribution, indicating most low-flow observations but a few high-flow days. Menstrual phase is narrow and tall, indicating consistently higher flow intensity.


Average Flow Volume Across Phases

Summarize average flow volume to observe general trends across the menstrual cycle.
```{r, echo=FALSE}
cycle_avg <- hormone_symptoms_eda |>
  group_by(phase) |>
  summarise(
    avg_flow_volume = mean(flow_volume_score, na.rm = TRUE),
    avg_appetite    = mean(appetite_score, na.rm = TRUE),
    .groups = "drop"
  )

# Flow Volume
ggplot(cycle_avg, aes(x = phase, y = avg_flow_volume, group = 1)) +
  geom_line(size = 1.2, color = "blue") +
  geom_point(size = 3, color = "red") +
  geom_text(aes(label = round(avg_flow_volume, 2)), vjust = -1, size = 4, fontface = "bold") +
  labs(title = "Average Flow Volume Across Phases",
       x = "Phase", y = "Flow Volume Score") +
  theme_minimal()+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    plot.title = element_text(face = "bold"),
    panel.grid.major = element_line(color = "grey90")
  )

```
Menstrual phase has the highest average flow (2.38), luteal is lowest (0.03), fertility is 0.07, and follicular is 0.23.

### Appetite

Average Appetite Across Phases

Summarize average appetite to observe general trends across the menstrual cycle.
```{r, echo=FALSE}
# Appetite
ggplot(cycle_avg, aes(x = phase, y = avg_appetite, group = 1)) +
  geom_line(size = 1.2, color = "purple") +
  geom_point(size = 3, color = "red") +
  geom_text(aes(label = round(avg_appetite, 2)), vjust = -1, size = 4, fontface = "bold") +
  labs(title = "Average Appetite Across Phases",
       x = "Phase", y = "Appetite Score") +
  theme_minimal()+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    plot.title = element_text(face = "bold"),
    panel.grid.major = element_line(color = "grey90")
  )
```
Appetite is highest in Fertility (2.96), lowest in Follicular (2.85), Luteal 2.95, Menstrual 2.90.




# Modelling


modeling from yue
### Mood SwingLinear Mixed Models

Test whether mood swing and fatigue differ by menstrual phase, accounting for repeated measures.
```{r, echo=FALSE}
mood_mixed_model <- lmer(moodswing_score ~ phase + (1 | id), data = hormone_symptoms_eda)


mood_mixed_model
```
Mood swings are lowest near ovulation, slightly higher in Luteal, and peak during Menstrual phase. Fatigue shows a similar pattern: lowest near ovulation, increasing through Luteal, highest during Menstrual phase. Random effects indicate that some individuals are chronically more fatigued or mood-sensitive than others, while residual variance captures day-to-day variability.

### Linear Mixed Model: Appetite vs Estrogen
Explore the relationship between estrogen levels and appetite scores, accounting for repeated measures.
```{r, echo=FALSE}
appetite_model <- lmer(appetite_score ~ estrogen + (1 | id), data = hormone_symptoms_eda)
summary(appetite_model)
```
Random intercept variance (0.1746) indicates participant-specific baseline appetite differences. The fixed effect of estrogen is small (0.00013 per unit) and not significant (t = 0.869).



# Discussion

### Data Handling
- data quality issues - take hormones & self-reported as an example

regardless the most common missing data (due to self-report), 1) integer values in categorical variables; 2) inconsistent likert scale ('very low/little' vs. 'very low') >> give us the tip when we run a study to collect different likert scale data, collect them at the same scale for easy and consistent future data analyses.

- data entry issues - take sleep.csv as an example

Yue (The observed hormonal, emotional, and physiological patterns across the menstrual cycle are highly consistent with established reproductive endocrinology. Estrogen gradually increases through the follicular phase, peaks during the fertility (ovulatory) phase, and remains at moderate levels in the luteal phase, while luteinizing hormone (LH) exhibits a sharp surge at ovulation to trigger follicle rupture and ovum release. Both hormones reach their lowest levels during menstruation, reflecting the endocrine “reset” following corpus luteum regression. Corresponding to these hormonal dynamics, mood swings show clear phase-dependent variation. The distribution of mood swing severity is most pronounced during menstruation, with average intensity highest in the menstrual phase (1.85) and lowest in the luteal phase (1.41), with intermediate levels in the follicular and fertility phases. This elevated emotional symptom burden during menstruation is biologically plausible, as the abrupt withdrawal of both estrogen and progesterone is known to disrupt serotonergic signaling and emotional regulation. Linear mixed-effects models further confirm significant phase-dependent effects on both mood swings and fatigue, with the largest increases occurring during menstruation and more moderate elevations during the luteal phase. The substantial random intercept variance indicates strong inter-individual differences in baseline symptom severity, reinforcing the necessity of accounting for repeated measures within participants.

Flow volume patterns also followed classic physiological expectations. The menstrual phase exhibited the highest and most consistent flow intensity (mean = 2.38), reflecting active endometrial shedding. In contrast, the luteal phase showed minimal flow (mean = 0.03) with a right-skewed distribution, indicating that most observations reflect very low flow with only occasional spotting. Fertility and follicular phases also remained low (0.07 and 0.23, respectively), consistent with the uterine lining being preserved outside of menstruation. These results strongly support the biological validity of the flow measurements.

Appetite demonstrated more subtle cycle-related variation, with the highest average levels occurring during the fertility phase (2.96), when estrogen is maximized, and the lowest during the follicular phase (2.85), with relatively stable values during luteal (2.95) and menstrual (2.90) phases. This suggests that appetite may be modestly influenced by hormonal elevation near ovulation. However, linear mixed-effects modeling revealed that estrogen alone was not a statistically significant predictor of appetite after accounting for repeated measures, with a very small effect size and a non-significant t-value (0.869). The random intercept variance indicates that baseline individual differences in appetite explain more variability than short-term hormonal fluctuations. Together, these findings suggest that while appetite may exhibit weak phase-related trends, it is regulated by a complex interplay of hormonal, behavioral, and individual-specific factors, and may be better modeled using multiple hormonal predictors such as LHalongside phase.)


## Limitation

Yue(Several limitations should be considered when interpreting these findings. First, the dataset relies on self-reported symptom scores, which may be subject to recall bias or subjective interpretation. Second, while repeated measures are accounted for using linear mixed models, the sample size for some phases may be limited, reducing statistical power to detect subtle effects. Finally, other biological and lifestyle factors—such as stress, sleep, diet, and physical activity—were not controlled for, which could influence mood, appetite, and fatigue. Therefore, the results should be interpreted as suggestive of phase-related trends rather than definitive causal effects.)

