---
title: "Report"
---

```{r, include=FALSE}
# Load packages 
library(tidyverse)
library(ggplot2)
library(modelr)
library(mgcv)
library(dplyr)
library(tidyr)
library(patchwork)
library(lme4)
library(lmerTest)

# Set plot theme
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.discrete.colour = function(...) scale_colour_brewer(..., palette = "Pastel2"),
  ggplot2.discrete.fill   = function(...) scale_fill_brewer(..., palette = "Pastel2"),

  ggplot2.continuous.colour = function(...) scale_colour_distiller(..., palette = "PuRd"),
  ggplot2.continuous.fill   = function(...) scale_fill_distiller(..., palette = "PuRd")
)
```

# Motivation

# Related Work

# Questions

# Data

## Data Sources

We utilized the inclusive, multimodal, longitudinal, and de-identified [mcPHASES](https://physionet.org/content/mcphases/1.0.0/) (menstrual cycle Physiological, Hormonal, and Self-Reported Events and Symptoms) dataset for menstrual health tracking with wearable devices.

Data was collected from 42 menstruating young adults in Canada over two 3-month intervals (one in 2022, and the other in 2024). Participants wore Fitbit Sense smartwatches to measure physiological signals, and Mira Plus Starter Kits to track their hormone levels. Participants also self-reported daily symptom experiences like cramps, sleep quality, and stress levels via a smartphone diary app.

The original dataset contains 23 structured tables organized by signal category. For this project, we focused on the lifestyle factors such as exercise, sleep, and stress, and menstrual cycle symptoms. The datasets we used are the following:

-   Demographics:
    -   `subject_info.csv`: static information such as age, education, and age at menarche
    -   `height_and_weight.csv`: participant's baselien height and weight
-   Hormone and Symptoms:
    -   `hormones_and_selfreport.csv`：self-reported daily data including symptoms (e.g., cramps, mood, menstrual flow) and manually entered hormone levels (LH, E3G, PdG)
-   Lifestyle Factors:
    -   `stress_score.csv` [TODO]
    -   `sleep.csv`: sleep stages, time in bed
    -   `sleep_score.csv`

------------------------------------------------------------------------

In total, we have 4 dataframes:

1.  `subject_info` (demographics data)
2.  `hormone_symptoms` (hormone and symtoms data)
3.  `sleep_stress` (sleep and stress related data)
4.  `exercise` (exercise related data).

For different analyses, we merged the dataframes above by participant `id` as needed.

------------------------------------------------------------------------

```{r load-lib, include=FALSE}
# Load necessary libraries
library(tidyverse)
library(patchwork)
library(ggsci)
```

### `subject_info` - Subject-Relevant Data

The `subject_info` dataframe contains participants demographics data from `subject_info.csv` and `height_and_weight.csv`, as well as participants study enrollment year (this piece of information is from `hormones_and_selfreport.csv`).

It contains a total of 62 observations of 19 variables. Some key variables are the following:

-   `id`: participant's unique identifier
-   Age relevant variables: `birth_year` and `age_of_first_menarche`
-   Height and weight related variables: `avg_height`, `avg_weight`, `bmi`
-   Study enrollment related information: `study` (flag of which study the participant enrolled in, `1`=2022, `2`=2024, `3`=both), `year` (the year of study)
-   Categorical variables like `ethnicity`, `education`, `self_report_menstrual_health_literacy`, etc.

```{r ht-and-wt, include=FALSE}
ht_wt = read_csv("data/height_and_weight.csv")|> 
  janitor::clean_names() |> 
  # calculate mean height/weight
  rowwise() |>
  mutate(
    avg_height = mean(c(height_2022, height_2024), na.rm = TRUE),
    avg_weight = mean(c(weight_2022, weight_2024), na.rm = TRUE)
  ) |>
  ungroup()
```

```{r, include=FALSE}
# import csv file
hormones_and_selfreport = read_csv("data/hormones_and_selfreport.csv") |> 
  janitor::clean_names() 
# extract study_flag
study_flag <- hormones_and_selfreport  |> 
  distinct(id, study_interval) |> 
  mutate(flag = 1,
         study_interval = paste0("study_", study_interval)) |> 
  pivot_wider(
    names_from = study_interval,
    values_from = flag,
    values_fill = 0  
  )
```

```{r, include=FALSE}
# read in subject-info.csv
subject_info_raw = read_csv("data/subject-info.csv") |> 
  janitor::clean_names()

subject_info <- subject_info_raw  |> 
  # merging
  left_join(ht_wt, by = "id") |> 
  left_join(study_flag, by = "id") |>
  # add a study_flag
  mutate(
    study = case_when(
      study_2022 == 1 & study_2024 == 1 ~ 3L,  # both 2022 & 2024
      study_2022 == 1 & study_2024 == 0 ~ 1L,  
      study_2024 == 1 & study_2022 == 0 ~ 2L,  
      TRUE ~ NA_integer_
    )
  ) |> 
  # pivot_long for wt & ht
  pivot_longer(
    cols = c(weight_2022, weight_2024, height_2022, height_2024),
    names_to  = c(".value", "year"), 
    names_sep = "_"                 
  ) |> 
  rename(
    weight_yr = weight,
    height_yr = height
  ) |> 
  # keep only the relevant study year
  filter(
    (study == 1 & year == 2022) |
    (study == 2 & year == 2024) |
    (study == 3)
  ) |> 
  # calculate age
  mutate(
    year = as.numeric(year),
    birth_year = as.numeric(birth_year),
    age = year - birth_year
  ) |> 
  # get a mean age for each id
  group_by(id) |> 
  mutate(
    mean_age = mean(age, na.rm = TRUE)
  ) |> 
  ungroup() |> 
  # get BMI
  mutate(
    bmi = weight_yr / ( (height_yr / 100)^2 )
  )
```

### `hormone_symptoms` - Hormone & Self-reported Data

[TODO] add overview

```{r, include=FALSE}
hormone_symptoms <- 
  read_csv("data/hormones_and_selfreport.csv") |> 
  janitor::clean_names() |> 
  # factor varialbes
  mutate(
    phase = factor(phase, levels = c("Menstrual", "Follicular", "Fertility", "Luteal")),
    fatigue = factor(fatigue, 
                     levels = c("Low", "Moderate", "High", "Very High", "Not at all")),
    moodswing = factor(moodswing,
                       levels = c("Very Low/Little", "Low", "Moderate", "High","Very High","Not at all")),
    flow_volume = factor(flow_volume, 
                         levels = c("Not at all", "Spotting / Very Light","Light","Moderate", "Somewhat Heavy","Heavy", "Very Heavy")),
    flow_color = factor(flow_color,
                        levels = c("Not at all", "Pink","Bright Red", "Dark Brown / Dark Red", "Grey", "Black","Other")),
    appetite = factor(appetite,
                      levels = c("Very Low","Low",  "Moderate", "High", "Very High"))
  ) |> 
  # convert categorical variable values to numeric intensity scores
  mutate(
    moodswing_num = case_when(
      moodswing == "Not at all" ~0,
      moodswing == "Very Low/Little" ~ 1,
      moodswing == "Low" ~ 2,
      moodswing == "Moderate" ~ 3,
      moodswing == "High" ~ 4,
      moodswing == "Very High" ~ 5,
      TRUE ~ NA_real_
    ),
    fatigue_num = case_when(
      fatigue == "Not at all" ~0,
      fatigue == "Low" ~ 1,
      fatigue == "Moderate" ~ 2,
      fatigue == "High" ~ 3,
      fatigue == "Very High" ~ 4,
      TRUE ~ NA_real_
    ),
    flow_volume_num = case_when(
      flow_volume == "Not at all" ~ 0,
      flow_volume == "Spotting / Very Light" ~ 1,
      flow_volume == "Light" ~ 2,
      flow_volume == "Moderate" ~ 3,
      flow_volume == "Somewhat Heavy" ~ 4,
      flow_volume == "Heavy" ~ 5,
      flow_volume == "Very Heavy" ~ 6,
      TRUE ~ NA_real_
    ),
    flow_color_num = case_when(
      flow_color == "Not at all" ~ 0,
      flow_color == "Pink" ~ 1,
      flow_color == "Bright Red" ~ 2,
      flow_color == "Dark Brown / Dark Red" ~ 3,
      flow_color == "Grey" ~ 4,
      flow_color == "Black" ~ 5,
      flow_color == "Other" ~ 6,
      TRUE ~ NA_real_
    ),
    appetite_num = case_when(
      appetite == "Very Low" ~ 0,
      appetite == "Low" ~ 1,
      appetite == "Moderate" ~ 2,
      appetite == "High" ~ 3,
      appetite == "Very High" ~ 4,
      TRUE ~ NA_real_
    )
  )

```

# EDA




# Modelling

# Discussion

## Limitation
