---
title: "sleep_stress_EDA"
output: github_document
---

```{r setup , include = FALSE}
library(tidyverse)
library(ggplot2)
library(modelr)
library(mgcv)
library(dplyr)
library(tidyr)
library(patchwork)

set.seed(1)
#library(readxl)
#library(haven)
theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

Load the data 

```{r, include=FALSE}
subject     = read_csv("data/subject-info.csv") |>
  janitor::clean_names()
sleep_score = read_csv("data/sleep_score.csv") |>
  janitor::clean_names()
sleep       = read_csv("data/sleep.csv") |>
  janitor::clean_names()
stress_score = read_csv("data/stress_score.csv") |>
  janitor::clean_names()
hormone = read_csv("data/hormones_and_selfreport.csv") |>
  janitor::clean_names()
```


View the sleep time for different participants in each study interval
```{r}
# plot the mean timein bed in hour
sleep |>
  group_by(id, study_interval)|>
  summarize(mean_sleep = mean(timeinbed/60),
            sd_sleep = sd(timeinbed),
            mean_efficiency = mean(efficiency),
            sd_efficiency = sd(efficiency)) |>
  ggplot(aes(x = id, y = mean_sleep, color = id))+
  geom_point()+
  facet_grid(study_interval~.)
```

The average sleep time for each person is different each study interval and varied across participants.  In 2022, some participants slept less than 5 hours and some slept more than 8 hours. In 2024, most participants have better sleep length, 7-8 hours.

Combine the sleep and stress score

```{r}
# join two data sets, using inner join
sleep_score1 = sleep_score |>
  select(id,study_interval, day_in_study, overall_score, deep_sleep_in_minutes)

sleep1 = sleep|>
   select(id,study_interval, sleep_start_day_in_study, timeinbed) |>
   mutate(day_in_study = sleep_start_day_in_study) |>
  select(-sleep_start_day_in_study)
stress_score1 = stress_score |>
  filter(calculation_failed == "FALSE")|>
  select(id, study_interval, day_in_study, stress_score)|>
  distinct(id, study_interval, day_in_study, .keep_all = TRUE)

sleep_stress = inner_join(sleep_score1, stress_score1, join_by("id", "study_interval", "day_in_study"))
 
# check of na values
sum(is.na(sleep_stress$overall_score))
sum(is.na(sleep_stress$stress_score))
```

Showing the sleep score and stress score changes 

```{r}
# graphs to show the score across ids
sleep_stress|>
  group_by(id) |> 
  summarize(
    mean_sleep_score = mean(overall_score, na.rm = TRUE),
    mean_stress_score = mean(stress_score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = id, mean_sleep_score)) +
  geom_smooth(aes(y = mean_sleep_score, color = "mean_sleep_score"), se= FALSE )+
  geom_smooth(aes(y = mean_stress_score, color = "mean_stress_score"), se = FALSE) 

```

Load and clean the hormone data for menstrual phases
```{r}
hormone1 = hormone |>
  select(id,phase, day_in_study, lh, estrogen,fatigue,moodswing)|>
  mutate(
    phase = factor(phase, levels = c("Follicular", "Fertility", "Luteal", "Menstrual")),
    fatigue = factor(fatigue, 
                     levels = c("Low", "Moderate", "High", "Very High")),
    moodswing = factor(moodswing,
                       levels = c("Very Low/Little", "Low", "Moderate", "High","Very High","Not at all"))
  )
phase_order <- c("Menstrual", "Follicular", "Fertility", "Luteal")

```

Joined data with the phase in hormone dataset

```{r}
sleep_stress_hormone =
  inner_join(sleep_stress,hormone1, join_by(id, day_in_study))
```

Now explore the changes of the stress score with the menstrual cycle. The participants slept best during the menstrual time, probably a good sign to rest well during this time. The lowest sleep score were recorded during ovulation/ fertility time in which the increased estrogen and lh levels may contribute to this low sleep quality. 
The higher stress score the less stressful people experience. In this study, the participants showed the least stressful during folicular period , as the bodies are full of energy. The most stressful time is during luteal period. 


```{r}

plot_1 = 
  sleep_stress_hormone |>
  group_by(phase) |>
  summarize(
    mean_sleep_score = mean(overall_score, na.rm = TRUE),
    mean_stress_score = mean(stress_score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = starts_with("mean_"),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(
    metric = recode(metric,
                    mean_sleep_score  = "Sleep Score",
                    mean_stress_score = "Stress Score"),
    phase = factor(phase, levels = phase_order)
  ) |>
  ggplot(aes(x = phase, y = value, color = metric, group = metric)) +
    geom_smooth(linewidth= 1) +
    geom_point(size = 3) +
    labs(color = "Metric", y = "Mean Score", x = "Phase") +
    labs(
    x = "Menstrual Cycle Phase",
    y = "Score"
  ) +
    theme_minimal() +
  theme(legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1))

plot_2  =
  sleep_stress_hormone |>
  group_by(phase) |>
  summarize(
    mean_lh = mean(lh, na.rm = TRUE),
    mean_estrogen = mean( estrogen, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = starts_with("mean_"),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(
    metric = recode(metric,
                    mean_lh           = "LH",
                    mean_estrogen     = "Estrogen"),
    phase = factor(phase, levels = phase_order)
  ) |>
  ggplot(aes(x = phase, y = value, color = metric, group = metric)) +
    geom_smooth(linewidth= 1) +
    geom_point(size = 3) +
    labs(color = "Metric", y = "level", x = "Phase") +
    labs(
    x = "Menstrual Cycle Phase",
    y = "Level"
  ) +
    theme_minimal() +
  theme(legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1))

plot_1+ plot_2
```

Variation of sleep score across menstrual cycle phases

```{r}
sleep_stress_hormone |>
  ggplot(aes(x = phase, y = overall_score, fill = phase))+
  geom_violin(alpha = 0.5)+
  geom_boxplot(width = 0.2)+
      labs(
    title = "Sleep Score Across Menstrual Cycle Phases",
    x = "Menstrual Cycle Phase",
    y = "score"
  ) +
    theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


Variation of stress score across menstrual cycle phases
```{r}
sleep_stress_hormone |>
  ggplot(aes(x = phase, y = stress_score, fill = phase))+
  geom_violin(alpha = 0.5)+
  geom_boxplot(width = 0.2)+
      labs(
    title = "Stress Score Across Menstrual Cycle Phases",
    x = "Menstrual Cycle Phase",
    y = "score"
  ) +
    theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```





Explore the variation among different participants. adding the id, ethnicity and birthyear of the subjects.

```{r}

sleep_stress_hormone2 = subject |>
  select(id, birth_year, ethnicity) |>
  mutate(age = 2025- birth_year) |>
  full_join(sleep_stress_hormone, join_by(id)) 

```

View the stress and sleep score across menstrual cycle phases for each 

```{r}
sleep_stress_hormone2 |>
  group_by(phase, study_interval, ethnicity) |>
  summarize(
    mean_sleep_score = mean(overall_score, na.rm = TRUE),
    mean_stress_score = mean(stress_score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = starts_with("mean_"),
    names_to = "metric",
    values_to = "value"
  ) |>
  drop_na()|>
  mutate(
    metric = recode(metric,
                    mean_sleep_score  = "Sleep Score",
                    mean_stress_score = "Stress Score"),
    phase = factor(phase, levels = phase_order)
  ) |>
  ggplot(aes(x = phase, y = value, color = ethnicity, group = ethnicity)) +
    geom_line(size = 1) +
    geom_point(size = 3) +
    facet_grid(study_interval ~ metric) +
    labs(color = "Ethnicity", y = "Mean Score", x = "Phase") +
    labs(
    title = "stress and sleep score Across Menstrual Cycle Phases",
    x = "Menstrual Cycle Phase",
    y = "score"
  ) +
    theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
 variation of sleep score and stress score across different ethnicity
 

```{r}
sleep_stress_hormone2 |>
  select(id, ethnicity, stress_score, overall_score) |>
  drop_na()|>
  pivot_longer(
    cols = c(stress_score, overall_score),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(
    metric = recode(metric,
      overall_score = "Sleep Score",
      stress_score  = "Stress Score"
    )
  ) |>
  ggplot(aes(x = ethnicity, y = value, fill = metric)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(y = "Score", fill = "Metric")

```

```{r}
library(lme4)

#  model
m1 <- lmer(stress_score ~ phase + study_interval + (1 | id), data = sleep_stress_hormone2)
summary(m1)

```



```{r}
sleep_stress_hormone_add =
  sleep_score1 |>
    inner_join(hormone1,      by = join_by(id, day_in_study)) |>
    inner_join(sleep1,        by = join_by(id, study_interval, day_in_study)) |>
    inner_join(stress_score1, by = join_by(id, study_interval, day_in_study))

# create a table
sleep_stress_hormone_add |>
  group_by(phase, id) |>
  summarize(
    timeinbed = mean(timeinbed/60)
  )|>
  pivot_wider(names_from = phase,
              values_from = timeinbed) |>
  knitr::kable(title = "average time in bed", digits = 1)
```


```{r}
sleep_stress_hormone_add |>
  group_by(phase, id) |>
  summarize(
    timeinbed = mean(timeinbed/60)
  )|> 
  ggplot(aes(x = phase, y = timeinbed, fill = phase)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.2) +
  labs(
    title = "Sleep time Across Menstrual Cycle Phases",
    x = "Menstrual Cycle Phase",
    y = "Time in bed (hrs)"
  ) +
    theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

See the correlation in sleep score file
```{r}
sleep_score |>
  group_by(id)|>
  summarise(mean_deep_sleep = mean(deep_sleep_in_minutes),
            mean_overall_score = mean(overall_score))|>
  ggplot(aes(x= mean_overall_score, y = mean_deep_sleep )) +
  geom_point()
```


NOT WORKING_try to put all in one scale
```{r, include= FALSE}

sleep_stress_hormone_plot = 
  sleep_stress_hormone |>
  group_by(phase) |>
  summarize(
    mean_sleep_score = mean(overall_score, na.rm = TRUE),
    mean_stress_score = mean(stress_score, na.rm = TRUE),
    mean_lh = mean(lh, na.rm = TRUE),
    mean_estrogen = mean( estrogen, na.rm = TRUE),
    .groups = "drop"
  ) 
### create new vars
left_vars  <- c("mean_sleep_score", "mean_stress_score")
right_vars <- c("mean_lh")
### scale
scaleFactor <- max(sleep_stress_hormone_plot[, right_vars], na.rm = TRUE)/
               max(sleep_stress_hormone_plot[, left_vars],  na.rm = TRUE)
scaleFactor2 = max(sleep_stress_hormone_plot[,"mean_estrogen"])/ max(sleep_stress_hormone_plot[,"mean_stress_score"])
## create plot
sleep_stress_hormone_plot|>
   pivot_longer(cols = starts_with("mean"),
               names_to = "metric",
               values_to = "value")|>
    mutate(phase = factor(phase, levels = phase_order),
    value_plot = case_when(metric == "mean_lh" ~ value / scaleFactor,  # scale lh up
                           metric == "mean_estrogen" ~ value/scaleFactor2,
                           metric %in% c("mean_sleep_score", "mean_stress_score") ~ value -30)
    )|>
  ggplot(aes(x = phase))+
  geom_smooth(aes(y = value_plot, color = metric, group = metric), size = 1) +
    scale_y_continuous(
    name = "Sleep & Stress Score (left axis)",
    #limits = c(65, 80),                #range of sleep score
    sec.axis = sec_axis( ~ . * scaleFactor,
  name = "Hormone Levels (LH & Estrogen)",
  breaks = c(0, 10, 50, 100))) +
  scale_color_manual(
    values = c(
      "mean_sleep_score" = "blue",
      "mean_stress_score" = "darkgreen",
      "mean_lh" = "red",
      "mean_estrogen" = "purple"
    )
  ) +
  theme_minimal(base_size = 14) 

```


See time in bed and deep sleep vs menstrual cycle

```{r}
sleep_stress_hormone_add |>
  group_by(phase)|>
   summarize(timeinbed = mean(timeinbed/60, na.rm = TRUE),
             deep_sleep_in_minutes = mean(deep_sleep_in_minutes/60, na.rm = TRUE)
             )|>
  ggplot(aes(x= phase)) +
  geom_line(aes( y = deep_sleep_in_minutes, group = 1),color = "blue",  linewidth = 0.8) +
  geom_line(aes(y = timeinbed, group = 1), color = "red", linewidth = 0.8)
```

