---
title: "Participants Characteristics"
author: "Muying Li"
date: "2025-11-18"
output: html_document
---

```{r load-lib, results='hide'}
library(tidyverse)
library(patchwork)
library(ggsci)
library(dplyr)
library(janitor)
library(gt)
```

Set plot theme
```{r}
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.discrete.colour = function(...) scale_colour_brewer(..., palette = "Pastel2"),
  ggplot2.discrete.fill   = function(...) scale_fill_brewer(..., palette = "Pastel2"),

  ggplot2.continuous.colour = function(...) scale_colour_distiller(..., palette = "PuRd"),
  ggplot2.continuous.fill   = function(...) scale_fill_distiller(..., palette = "PuRd")
)
```

# Data Preparation

### `ht_wt` - Subject height and weight 
- `id`
- `height`: mean height
- `weight`: mean weight
- `enrollment_year`: some participants were enrolled in both 2022 and 2024, keep the latest one as enrollment year
- `height_2022`
- `height_2024`
- `weight_2022`
- `weight_2024`

Process:

1. Take the mean for subjects who have multiple heights and weights (both in 2022 and 2024), take the mean.
2. Note missing data
```{r ht-and-wt}
ht_wt = read_csv("data/height_and_weight.csv")|> 
  janitor::clean_names() |> 
  # calculate mean height/weight
  rowwise() |>
  mutate(
    avg_height = mean(c(height_2022, height_2024), na.rm = TRUE),
    avg_weight = mean(c(weight_2022, weight_2024), na.rm = TRUE)
  ) |>
  ungroup()
```

Count how many missing values:
```{r missing-value-ht-wt}
ht_wt |>
  summarise(
    missing_ht = sum(is.na(avg_height)),
    missing_wt = sum(is.na(avg_weight)),
    missing_both = sum(is.na(avg_weight) & is.na(avg_height))
  )

ht_wt |>
  filter(is.na(avg_height) | is.na(avg_weight))
```

### `hormones_and_selfreport` - hormone and self-reported data
```{r}
hormones_and_selfreport = read_csv("data/hormones_and_selfreport.csv") |> 
  janitor::clean_names() 
```

Get participants study enrollment info:

- `id`
- `study_2022`: whether enrolled in 2022 (1=Yes, 0=No)
- `study_2024`: whether enrolled in 2024 (1=Yes, 0=No)
```{r}
study_flag <- hormones_and_selfreport  |> 
  distinct(id, study_interval) |> 
  mutate(flag = 1,
         study_interval = paste0("study_", study_interval)) |> 
  pivot_wider(
    names_from = study_interval,
    values_from = flag,
    values_fill = 0  
  )
```

### `subject_info` - subject information (all demographic data including BMI)
```{r}
# read in csv
subject_info_raw = read_csv("data/subject-info.csv") |> 
  janitor::clean_names()
```

Process:
1. merge with `ht_wt` and `study_flag`
2. `pivot_long` for ht, wt, and study (1=2022, 2=2024, 3=2025)
3. calculate age based on enrollment study year
4. calculate BMI
```{r}
subject_info <- subject_info_raw  |> 
  # merging
  left_join(ht_wt, by = "id") |> 
  left_join(study_flag, by = "id") |>
  # add a study_flag
  mutate(
    study = case_when(
      study_2022 == 1 & study_2024 == 1 ~ 3L,  # both 2022 & 2024
      study_2022 == 1 & study_2024 == 0 ~ 1L,  
      study_2024 == 1 & study_2022 == 0 ~ 2L,  
      TRUE ~ NA_integer_
    )
  ) |> 
  # pivot_long for wt & ht
  pivot_longer(
    cols = c(weight_2022, weight_2024, height_2022, height_2024),
    names_to  = c(".value", "year"), 
    names_sep = "_"                 
  ) |> 
  rename(
    weight_yr = weight,
    height_yr = height
  ) |> 
  # keep only the relevant study year
  filter(
    (study == 1 & year == 2022) |
    (study == 2 & year == 2024) |
    (study == 3)
  ) |> 
  # calculate age
  mutate(
    year = as.numeric(year),
    birth_year = as.numeric(birth_year),
    age = year - birth_year
  ) |> 
  # get a mean age for each id
  group_by(id) |> 
  mutate(
    mean_age = mean(age, na.rm = TRUE)
  ) |> 
  ungroup() |> 
  # get BMI
  mutate(
    bmi = weight_yr / ( (height_yr / 100)^2 )
  ) |>
  #this factors variables and also creates "gender abrv" and "education abrv", which are simplified versions of these variables for cleaner tables and figures 
  mutate(
    education = str_trim(education),
    education_abrv = case_when(
      education == "High school degree or equivalent (e.g. GED)" ~ "High School",
      education == "Some university/ post-secondary, no degree" ~ "Some College",
      education == "Bachelor's degree (e.g. BA, BS)" ~ "Bachelors",
      education %in% c("Master's degree (e.g. MA, MS, MEd)", "Doctorate or professional degree") ~ "Graduate Degree or Higher",
      TRUE ~ "Other"),
    education_abrv = factor(
      education_abrv,
      levels = c("High School", "Some College", "Bachelors", "Graduate Degree or Higher", "Other"),
      ordered = TRUE),
     self_report_menstrual_health_literacy = factor(
      self_report_menstrual_health_literacy,
      levels = c("Non-existant", "Low", "Medium", "High"),
      ordered = TRUE),
    gender_abrv = case_when(
      gender %in% c("Gender Fluid", "Non-binary") ~ "Gender Fluid or Non-binary", 
      gender %in% c("Prefer not to say", "Other") ~ "Prefer not to say or Other", 
      gender == "Woman" ~ "Woman"),
      gender_abrv = factor(gender_abrv, levels = c("Woman", "Gender Fluid or Non-binary", "Prefer not to say or Other")), 
      sexually_active = factor(sexually_active, levels = c("Yes", "No", "Prefer not to say")))
```

# Report Summary

### Summary Table 
```{r}
#This is setup to make a clean table with simplified variables, count, percent count, except for menarche which is mean and (range)
#Menarche
age_row = subject_info %>%
  summarise(variable = "Age at Menarche",
    subgroup = "",
    value = sprintf("%.1f (%d–%d)",
      mean(age_of_first_menarche, na.rm = TRUE),
      min(age_of_first_menarche, na.rm = TRUE),
      max(age_of_first_menarche, na.rm = TRUE)))

#bmi
bmi_row = subject_info %>%
  summarise(variable = "BMI",
    subgroup = "",
    value = sprintf("%.1f (%.1f-%.1f)",
      mean(bmi, na.rm = TRUE),
      min(bmi, na.rm = TRUE),
      max(bmi, na.rm = TRUE)))

#Race
ethnicity_rows = subject_info %>%
  count(ethnicity) %>%
  mutate(variable = "Ethnicity",
    subgroup = ethnicity,
    value = sprintf("%d (%.1f%%)", n, 100 * n / sum(n))) %>%
  select(variable, subgroup, value)

#Gender
gender_rows = subject_info %>%
  count(gender_abrv) %>%
  mutate(variable = "Gender",
    subgroup = gender_abrv,
    value = sprintf("%d (%.1f%%)", n, 100 * n / sum(n))) %>%
  select(variable, subgroup, value)

#Sexually Active 
sex_rows = subject_info %>%
  count(sexually_active) %>%
  mutate(variable = "Sexually Active",
    subgroup = sexually_active,
    value = sprintf("%d (%.1f%%)", n, 100 * n / sum(n))) %>%
  select(variable, subgroup, value)

#Education 
education_rows = subject_info %>%
  count(education_abrv) %>%
  mutate(variable = "Education", 
         subgroup = education_abrv,
         value = sprintf("%d (%.1f%%)", n, 100 * n / sum(n))) %>%
  select(variable, subgroup, value)

#Combined
summary_table = bind_rows(
  age_row,
  bmi_row,
  ethnicity_rows,
  gender_rows,
  education_rows,
  sex_rows)

summary_table_clean <- summary_table %>%
  group_by(variable) %>%
  mutate(variable = ifelse(row_number() == 1, variable, "")) %>%
  ungroup()

summary_table_gt = summary_table_clean %>%
  gt() %>%
  cols_label(variable = "",
    subgroup = "",
    value = "Mean (Max-Min) or Count (%)") %>%
  tab_options(table.font.size = px(14),
    data_row.padding = px(6)) %>%
  tab_style(style = cell_text(weight = "bold"),
    locations = cells_body(columns = variable)) %>%
cols_align(align = "right",
    columns = value) 

summary_table_gt
```
# Visualization
### Gender
```{r}
subject_info |> 
  # factor gender
  mutate(
    gender = factor(
      gender,
      levels=c("Woman","Non-binary","Gender Fluid","Other","Prefer not to say")
    )
  ) |> 
  ggplot(aes(gender, fill=gender)) + 
  geom_bar() +
  labs(
    title = "Gender Distribution",
    x = "Gender Category",
    y = "Count"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none" 
    )
```

### Ethnicity
```{r}
ggplot(subject_info, aes(fct_infreq(ethnicity), ,fill=fct_infreq(ethnicity))) + 
  geom_bar()+
  labs(
    title = "Ethnicity Distribution",
    x = "Ethnicity",
    y = "Count"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none" 
    )
```

### Education 
```{r}
subject_info |>  
  # factor education level
  mutate(
    education = factor(
      education,
      levels=c("High school degree or equivalent (e.g. GED)","Some university/ post-secondary, no degree","Bachelor's degree (e.g. BA, BS)"    ,"Master's degree (e.g. MA, MS, MEd)","Doctorate or professional degree")
    )
  ) |> 
  ggplot(aes(education, fill=education)) + 
  geom_bar() +
  coord_flip()+
  labs(
    title = "Education Level Distribution",
    x = "Education Level",
    y = "Count"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none" 
    )

```

### Sexually active
```{r}
subject_info |> 
  mutate(sexually_active = fct_inorder(sexually_active)) |> 
  ggplot(aes(x = sexually_active, fill = sexually_active)) +
  geom_bar() +
  labs(
    title = "Sexual Activity Status",
    x = "Sexually Active",
    y = "Count"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none" 
    )
```

### Menstrual health literacy

ordered by level
```{r}
subject_info |> 
  filter(!is.na(self_report_menstrual_health_literacy)) |> 
  count(self_report_menstrual_health_literacy) |> 
  mutate(
    prop = n / sum(n),
    mhl = factor(
      self_report_menstrual_health_literacy,
      levels = c("Non-existent", "Low", "Medium", "High", "Expert")
    )  
  ) |> 
  ggplot(aes(x = mhl,
             y = n,
             fill = mhl)) +
  geom_col() +
  coord_flip() +   # horizontal for readability
  labs(
    title = "Self-Reported Menstrual Health Literacy",
    x = "Menstrual Health Literacy Level",
    y = "Count"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none"
  )

education_health_lit = ggplot(subject_info, aes(x = education_abrv, fill = self_report_menstrual_health_literacy)) +
  geom_bar(position = "fill") + 
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Self-Reported Menstrual Health Literacy by Education Level",
    x = "Education Level",
    y = "Percentage",
    fill = "Menstrual Health Literacy") +
  theme_minimal(base_size = 14)

education_health_lit
```


### Age distribution

- View mean age in total (2022 & 2024)
- View age by study (2022, 2024)
```{r age-distribution-boxplot}
overall_n <- subject_info |> 
  distinct(id) |> 
  count(name = "n")

age_boxplot_all <- subject_info |> 
  ggplot(aes(x="All participants", y = mean_age)) +
  geom_boxplot(fill = "#bcbddc")+
  geom_text(
    data = overall_n,
    aes(
      x = "All participants", 
      y = max(subject_info$age, na.rm = TRUE) + 3,
      label = paste0("n = ", n)
    )
  ) +
  labs(
    x = NULL,
    y = "Mean age (years)",
    title = "Mean age of all participants"
  ) +
  ylim(15, NA) +
  theme(
    axis.text.x = element_text(size = 10),
    plot.title  = element_text(hjust = 0.5)
  )

count_df <- subject_info |> 
  distinct(id, year) |> 
  count(year, name="n")

age_boxplot_by_study <- subject_info |> 
  ggplot(aes(x = factor(year), y = age, fill = factor(year))) +
  geom_boxplot() +
  geom_text(
    data = count_df,
    aes(x = factor(year), y = max(subject_info$age, na.rm = TRUE) + 1, 
        label = paste0("n = ", n)),
    size = 4
  ) +
  labs(
    x = "Year",
    y = "Age (years)",
    fill = "Year",
    title = "Age Distribution by Study Year"
  ) +
  ylim(15, NA) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none" 
  )

age_boxplots <- age_boxplot_all + age_boxplot_by_study
```
```{r}
age_boxplots
```

### Age of first menarche
```{r}
subject_info |> 
  distinct(id, .keep_all = TRUE) |> 
  filter(!is.na(age_of_first_menarche)) |> 
  ggplot(aes(x = age_of_first_menarche, fill = factor(age_of_first_menarche))) +
  geom_bar() +
  labs(
    title = "Distribution of Age at First Menarche",
    x = "Age at First Menarche (years)",
    y = "Count"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.title = element_text(size = 12),
    legend.position = "none"
  )

menarche_plot = ggplot(subject_info, aes(x = "", y = age_of_first_menarche)) +
  geom_boxplot(fill = "cornflowerblue") +
  labs(title = "Age of Menarche",
    y = "Age (years)",
    x = "") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_blank(),
    axis.ticks.x = element_blank())

menarche_plot
```

Findings: most common age of 1st menarche is at 12 y.o.

## Anthropometrics
### Weight, Height, BMI distribution

- boxplot for weight (all (mean)vs. by study)
```{r}
overall_weight_n <- subject_info |> 
  filter(!is.na(weight_yr)) |> 
  summarize(n = n())

weight_all <- subject_info |> 
  filter(!is.na(weight_yr)) |> 
  ggplot(aes(x = "All participants", y = weight_yr)) +
  geom_boxplot(fill = "#bcbddc") +
  geom_text(
    data = overall_weight_n,
    aes(
      x = "All participants",
      y = 135, 
      label = paste0("n = ", n)
    )
  ) +
  labs(
    x = NULL,
    y = "Weight (kg)",
    title = "Overall Mean Weight Distribution"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(size = 10)
  )

weight_year_n <- subject_info |> 
  filter(!is.na(weight_yr)) |> 
  count(year, name = "n")

weight_by_study <- subject_info |> 
  filter(!is.na(weight_yr)) |> 
  ggplot(aes(x = factor(year), y = weight_yr, fill = factor(year))) +
  geom_boxplot() +
  geom_text(
    data = weight_year_n,
    aes(
      x = factor(year),
      y = 135,
      label = paste0("n = ", n)
    )
  ) +
  labs(
    x = "Year",
    y = "Weight (kg)",
    fill = "Year",
    title = "Weight Distribution by Study Year"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none" 
  )

weight_boxplots <- weight_all + weight_by_study
```

```{r}
weight_boxplots
```

<span style="color: red;">Note an extreme weight value, need to exclude for the future weight-related analyses</span>

- boxplot for height (all (mean)vs. by study)
```{r}
overall_height_n <- subject_info |> 
  filter(!is.na(height_yr)) |> 
  summarize(n = n())

height_all <- subject_info |> 
  filter(!is.na(height_yr)) |> 
  ggplot(aes(x = "All participants", y = height_yr)) +
  geom_boxplot(fill = "#bcbddc") +
  geom_text(
    data = overall_height_n,
    aes(
      x = "All participants",
      y = 190, 
      label = paste0("n = ", n)
    )
  ) +
  labs(
    x = NULL,
    y = "Height (cm)",
    title = "Overall Mean Height Distribution"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(size = 10)
  )

height_year_n <- subject_info |> 
  filter(!is.na(height_yr)) |> 
  count(year, name = "n")

height_by_study <- subject_info |> 
  filter(!is.na(height_yr)) |> 
  ggplot(aes(x = factor(year), y = height_yr, fill = factor(year))) +
  geom_boxplot() +
  geom_text(
    data = height_year_n,
    aes(
      x = factor(year),
      y = 190,
      label = paste0("n = ", n)
    )
  ) +
  labs(
    x = "Year",
    y = "Height (cm)",
    fill = "Year",
    title = "Height Distribution by Study Year"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none" 
  )

height_boxplots <- height_all + height_by_study
```

```{r}
height_boxplots
```

No outlier noticed.

- boxplot for BMI
```{r bmi-boxplot}
overall_bmi_n <- subject_info |> 
  filter(!is.na(bmi)) |> 
  summarize(n = n())

bmi_all <- subject_info |> 
  filter(!is.na(bmi)) |> 
  ggplot(aes(x = "All participants", y = bmi)) +
  geom_boxplot(fill = "#bcbddc") +
  geom_text(
    data = overall_bmi_n,
    aes(
      x = "All participants",
      y = 45, 
      label = paste0("n = ", n)
    )
  ) +
  labs(
    x = NULL,
    y = "BMI (kg/m^2)",
    title = "Overall Mean BMI Distribution"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(size = 10)
  )
bmi_study_n <- subject_info |> 
  filter(!is.na(bmi)) |> 
  count(year, name = "n")

bmi_by_study <- subject_info |> 
  filter(!is.na(bmi)) |> 
  ggplot(aes(x = factor(year), y = bmi, fill = factor(year))) +
  geom_boxplot() +
  geom_text(
    data = bmi_study_n,
    aes(
      x = factor(year),
      y = 45,
      label = paste0("n = ", n)
    )
  ) +
  labs(
    x = "Year",
    y = "BMI (kg/m^2)",
    fill = "Year",
    title = "BMI Distribution by Study Year"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none" 
  )

bmi_boxplots <- bmi_all + bmi_by_study
```

```{r}
bmi_boxplots
```


